---
title: "Data606Lab2"
author: "Stephanie Roark"
date: "9/22/2018"
output:
  html_document:
    theme: cosmo
    highlight: pygments
---

## Hot Hands

Basketball players who make several baskets in succession are described as 
having a *hot hand*. Fans and players have long believed in the hot hand 
phenomenon, which refutes the assumption that each shot is independent of the 
next. However, a 1985 paper by Gilovich, Vallone, and Tversky collected evidence
that contradicted this belief and showed that successive shots are independent 
events ([http://psych.cornell.edu/sites/default/files/Gilo.Vallone.Tversky.pdf](http://psych.cornell.edu/sites/default/files/Gilo.Vallone.Tversky.pdf)). This paper started a great controversy that continues to this day, as you can 
see by Googling *hot hand basketball*.

We do not expect to resolve this controversy today. However, in this lab we'll 
apply one approach to answering questions like this. The goals for this lab are 
to (1) think about the effects of independent and dependent events, (2) learn 
how to simulate shooting streaks in R, and (3) to compare a simulation to actual
data in order to determine if the hot hand phenomenon appears to be real.

Our investigation will focus on the performance of one player: Kobe Bryant of 
the Los Angeles Lakers. His performance against the Orlando Magic in the 2009 
NBA finals earned him the title *Most Valuable Player* and many spectators 
commented on how he appeared to show a hot hand. Let's load some data from those
games and look at the first several rows.

```{r load-data, eval=TRUE}
load("more/kobe.RData")
head(kobe)
```

In this data frame, every row records a shot taken by Kobe Bryant. If he hit the
shot (made a basket), a hit, `H`, is recorded in the column named `basket`, 
otherwise a miss, `M`, is recorded.

Just looking at the string of hits and misses, it can be difficult to gauge 
whether or not it seems like Kobe was shooting with a hot hand. One way we can 
approach this is by considering the belief that hot hand shooters tend to go on 
shooting streaks. For this lab, we define the length of a shooting streak to be 
the *number of consecutive baskets made until a miss occurs*.

For example, in Game 1 Kobe had the following sequence of hits and misses from 
his nine shot attempts in the first quarter:

```{r first9, eval=TRUE}
kobe$basket[1:9]
```

Within the nine shot attempts, there are six streaks, which are separated by a 
"|" above. Their lengths are one, zero, two, zero, zero, zero (in order of 
occurrence).

    Exercise 1.  What does a streak length of 1 mean, i.e. how many hits and misses are in a streak of 1? What about a streak length of 0?

A streak of length 1 means that he made only 1 shot in the streak and then missed the next shot. A streak of length 0 would be a miss only with no shots made.


###Calculate the lengths of all shooting streaks and then look at the distribution:

```{r calc-streak-kobe, eval=TRUE}
kobe_streak <- calc_streak(kobe$basket)
barplot(table(kobe_streak))
```

    Exercise 2.  Describe the distribution of Kobe's streak lengths from the 2009 NBA finals. What was his typical streak length? How long was his longest streak of baskets?

The distribution is right skewed with the majority of his shot streaks being misses followed by just one shot made and missing the next. Therefore the typical streak length is 0 and the longest streak was 4 shots before missing.

## Compared to What?

We've shown that Kobe had some long shooting streaks, but are they long enough 
to support the belief that he had hot hands? What can we compare them to?

To answer these questions, let's return to the idea of *independence*. Two 
processes are independent if the outcome of one process doesn't effect the outcome 
of the second. If each shot that a player takes is an independent process, 
having made or missed your first shot will not affect the probability that you
will make or miss your second shot.

A shooter with a hot hand will have shots that are *not* independent of one 
another. Specifically, if the shooter makes his first shot, the hot hand model 
says he will have a *higher* probability of making his second shot.

Let's suppose for a moment that the hot hand model is valid for Kobe. During his
career, the percentage of time Kobe makes a basket (i.e. his shooting 
percentage) is about 45%, or in probability notation,

\[ P(\textrm{shot 1 = H}) = 0.45 \]

If he makes the first shot and has a hot hand (*not* independent shots), then 
the probability that he makes his second shot would go up to, let's say, 60%,

\[ P(\textrm{shot 2 = H} \, | \, \textrm{shot 1 = H}) = 0.60 \]

As a result of these increased probabilites, you'd expect Kobe to have longer 
streaks. Compare this to the skeptical perspective where Kobe does *not* have a
hot hand, where each shot is independent of the next. If he hit his first shot,
the probability that he makes the second is still 0.45.

\[ P(\textrm{shot 2 = H} \, | \, \textrm{shot 1 = H}) = 0.45 \]

In other words, making the first shot did nothing to effect the probability that
he'd make his second shot. If Kobe's shots are independent, then he'd have the 
same probability of hitting every shot regardless of his past shots: 45%.

Now that we've phrased the situation in terms of independent shots, let's return
to the question: how do we tell if Kobe's shooting streaks are long enough to 
indicate that he has hot hands? We can compare his streak lengths to someone
without hot hands: an independent shooter. 


##Simulations in R

If you wanted to simulate flipping a fair coin 100 times, you could either run 
the function 100 times or, more simply, adjust the `size` argument, which 
governs how many samples to draw (the `replace = TRUE` argument indicates we put
the slip of paper back in the hat before drawing again). 

```{r sim-fair-coin, eval=TRUE}
outcomes <- c("heads", "tails")
sample(outcomes, size = 1, replace = TRUE)
sim_fair_coin <- sample(outcomes, size = 100, replace = TRUE)
```


```{r table-sim-fair-coin, eval=TRUE}
sim_fair_coin
table(sim_fair_coin)
```

Since there are only two elements in `outcomes`, the probability that we "flip" 
a coin and it lands heads is 0.5. Say we're trying to simulate an unfair coin 
that we know only lands heads 20% of the time. We can adjust for this by adding 
an argument called `prob`, which provides a vector of two probability weights.

```{r sim-unfair-coin, eval=TRUE}
sim_unfair_coin <- sample(outcomes, size = 100, replace = TRUE, prob = c(0.2, 0.8))
```

`prob=c(0.2, 0.8)` indicates that for the two elements in the `outcomes` vector,
we want to select the first one, `heads`, with probability 0.2 and the second 
one, `tails` with probability 0.8. Another way of thinking about this is to 
think of the outcome space as a bag of 10 chips, where 2 chips are labeled 
"head" and 8 chips "tail". Therefore at each draw, the probability of drawing a 
chip that says "head"" is 20%, and "tail" is 80%.

    Exercise 3.  In your simulation of flipping the unfair coin 100 times, how many flips 
    came up heads?

```{r    eval=TRUE}
sum(sim_unfair_coin == "heads")
```

In a sense, we've shrunken the size of the slip of paper that says "heads", 
making it less likely to be drawn and we've increased the size of the slip of 
paper saying "tails", making it more likely to be drawn. When we simulated the 
fair coin, both slips of paper were the same size. 


## Simulating the Independent Shooter

Simulating a basketball player who has independent shots uses the same mechanism 
that we use to simulate a coin flip. To simulate a single shot from an 
independent shooter with a shooting percentage of 50% we type,

```{r sim-basket, eval=TRUE}
outcomes <- c("H", "M")
sim_basket <- sample(outcomes, size = 1, replace = TRUE)
sim_basket
```

To make a valid comparison between Kobe and our simulated independent shooter, 
we need to align both their shooting percentage and the number of attempted shots.

    Exercise 4.  What change needs to be made to the `sample` function so that it reflects a 
    shooting percentage of 45%? Make this adjustment, then run a simulation to 
    sample 133 shots. Assign the output of this simulation to a new object called
    `sim_basket`.

```{r sim-basket exercise, eval=TRUE}
sim_basket <- sample(outcomes, size = 133, replace = TRUE, prob = c(.45, .55))
table(sim_basket)
```

Note that we've named the new vector `sim_basket`, the same name that we gave to
the previous vector reflecting a shooting percentage of 50%. 

With the results of the simulation saved as `sim_basket`, we have the data 
necessary to compare Kobe to our independent shooter. We can look at Kobe's data 
alongside our simulated data.

```{r compare-basket, eval=TRUE}
kobe$basket
sim_basket
```

Both data sets represent the results of 133 shot attempts, each with the same 
shooting percentage of 45%. We know that our simulated data is from a shooter 
that has independent shots. That is, we know the simulated shooter does not have
a hot hand.

* * *

## On your own

### Comparing Kobe Bryant to the Independent Shooter

Using `calc_streak`, compute the streak lengths of `sim_basket`.

```{r  calc streak for independent shooter, eval=TRUE}
IndependentShooter_streak <- calc_streak(sim_basket)
barplot(table(IndependentShooter_streak))
```


-   Describe the distribution of streak lengths. What is the typical streak 
    length for this simulated independent shooter with a 45% shooting percentage?
    How long is the player's longest streak of baskets in 133 shots?


The distribution looks very similar to Kobe's streak barplot with the majority of shots being misses or streaks of 0 followed by streaks of 1 shot. The longest streak changes whenever a new simulation is run but for this simulation the longest streak is **`r I(max(IndependentShooter_streak))`**.



-   If you were to run the simulation of the independent shooter a second time, 
    how would you expect its streak distribution to compare to the distribution 
    from the question above? Exactly the same? Somewhat similar? Totally 
    different? Explain your reasoning.


```{r  calc streak for 2nd independent shooter, eval=TRUE}
sim_basket <- sample(outcomes, size = 133, replace = TRUE, prob = c(.45, .55))
IndependentShooter_streak <- calc_streak(sim_basket)
barplot(table(IndependentShooter_streak))
```


I would expect the new simulation to be somewhat similar but not exactly the same as the first distribution because the parameters for the simulation are the same but each new simulation will have a different result. The longest streak for this new simulation is **`r I(max(IndependentShooter_streak))`**.


-   How does Kobe Bryant's distribution of streak lengths compare to the 
    distribution of streak lengths for the simulated shooter? Using this 
    comparison, do you have evidence that the hot hand model fits Kobe's 
    shooting patterns? Explain.

Let's look again at Kobe's streak lengths:

```{r plot streak-kobe, eval=TRUE}
barplot(table(kobe_streak))
```

The Independent Shooter streak distribution is almost identical to Kobe's streak distribution. Both shooters have far more streaks of length 0 and the next longest streak is 1 for both. The similarity of the distributions clearly shows that Kobe is no more likely to have longer shooting streaks than the Independent Shooter. If Kobe's distribution fit the hot hands model then his streaks would be longer than the Independent Shooter's but the barplots clearly show that this is not the case. Therefore the Hot Hands hypothesis is rejected.