---
title: 'Data 606: Lab  0'
author: "Stephanie Roark"
date: "9/2/2018"
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cosmo
  pdf_document: default
---
#Comparing Arbutnot's Baptisms to Present Day Births

##Arbuthnot's Baptism Data
The Arbuthnot data set refers to Dr. John Arbuthnot, an 18<sup>th</sup> century 
physician, writer, and mathematician.  He was interested in the ratio of newborn
boys to newborn girls, so he gathered the baptism records for children born in
London for every year from 1629 to 1710.

A summary of Arbuthnot's baptism data set for girls and boys:

```{r view-data, eval=TRUE}
source("more/arbuthnot.R")
head(arbuthnot)
dim(arbuthnot)
names(arbuthnot)
```


#Present Day Birth Records in the US

The Present data set contains trend analysis of the sex ratios at birth in the US from 19049 to 2002.

##Question 1:
##What are the dimensions of the data frame and what are the variable or column names?

The present data set contains 63 rows and 3 columns which are named year, boys, and girls.

```{r view present data , eval=TRUE}
source("more/present.R")
head(present)
dim(present)
names(present)
```

##What years are included in this data set?

The  birth data ranges from 1940 to 2002.

```{r max and min of year, eval=TRUE}
min(present$year)

max(present$year)
```

##Question 2: How do these counts compare to Arbuthnot’s? 

##Total number of baptisms per year 

The total number of baptisms for boys and girls for the period from 1629 to 1710.

```{r plot total baptisms, eval=TRUE}
plot(arbuthnot$year, arbuthnot$boys + arbuthnot$girls, type = "l")
```

##Total number of births per year 

The total number of births for boys and girls for the period from 1940 to 2002.


```{r plot total birth per year, eval=TRUE}
plot(present$year, present$boys + present$girls, type = "l")
```

##Are they on a similar scale?
###Scale for Baptism data
```{r max & min for baptisms, eval=TRUE}
min(arbuthnot$boys + arbuthnot$girls)
max(arbuthnot$boys + arbuthnot$girls)
```

###Scale for modern US birth data
```{r max and min for births, eval=TRUE}
min(present$boys + present$girls)
max(present$boys + present$girls)
```

###Ratio of min and max births to baptisms 
The scale of present day births is several hundred times the scale of baptisms.
```{r comparing max and mins, eval=TRUE}
min(present$boys + present$girls)/min(arbuthnot$boys + arbuthnot$girls)
max(present$boys + present$girls)/max(arbuthnot$boys + arbuthnot$girls)
```

##Question 3: Make a plot that displays the boy-to-girl ratio for every year in the data set. What do you see?

Early on the number of boys born is higher than girls, but as the years progress the proportion of boys to girls gets closer to being equal. 


###Plot of ratio of boy to girl ratio
```{r plot present ration of boys to girls, eval=TRUE}
plot(present$year, present$boys/present$girls, type = "l")
```

###Proportion of boys to girls

There are more boys born than girls during the entire period of the data set, but the proportion of boys gets smaller over time.

```{r proportion of boys to girls, eval=TRUE}

present$boys/(present$boys + present$girls)
plot(present$year, present$boys/(present$boys + present$girls), type = "l")
```

##Question 4: In what year did we see the most total number of births in the U.S.?

The year with the greatest number of total births is 1961. 

```{r year with most births, eval=TRUE}
present$totalBirths = present$boys + present$girls
present[ present$totalBirths == max(present$totalBirths), ]
```