---
title: "Data 606 Lab1"
author: "Stephanie Roark"
date: "9/9/2018"
output:
 html_document:
    theme: cosmo
    highlight: pygments
---

## Intro to Data

## Getting started

The Behavioral Risk Factor Surveillance System (BRFSS) is an annual telephone 
survey of 350,000 people in the United States. As its name implies, the BRFSS 
is designed to identify risk factors in the adult population and report 
emerging health trends. For example, respondents are asked about their diet and 
weekly physical activity, their HIV/AIDS status, possible tobacco use, and even
their level of healthcare coverage. The BRFSS Web site 
([http://www.cdc.gov/brfss](http://www.cdc.gov/brfss)) contains a complete 
description of the survey, including the research questions that motivate the 
study and many interesting results derived from the data.

We will focus on a random sample of 20,000 people from the BRFSS survey 
conducted in 2000. While there are over 200  variables in this data set, we will
work with a small subset.

We begin by loading the data set of 20,000 observations into the R workspace. 
After launching RStudio, enter the following command.

```{r load-data, eval=TRUE}
source("more/cdc.R")
```


Some define Statistics as the field that focuses on turning information into
knowledge.  The first step in that process is to summarize and describe the raw
information - the data.  In this lab, you will gain insight into public health
by generating simple graphical and numerical summaries of a data set collected 
by the Centers for Disease Control and Prevention (CDC).  As this is a large 
data set, along the way you'll also learn the indispensable skills of data 
processing and subsetting.

```{r names, eval=TRUE}
names(cdc)
```

This returns the names `genhlth`, `exerany`, `hlthplan`, `smoke100`, `height`, 
`weight`, `wtdesire`, `age`, and `gender`. Each one of these variables 
corresponds to a question that was asked in the survey.  For example, for 
`genhlth`, respondents were asked to evaluate their general health, responding
either excellent, very good, good, fair or poor. The `exerany` variable 
indicates whether the respondent exercised in the past month (1) or did not (0).
Likewise, `hlthplan` indicates whether the respondent had some form of health 
coverage (1) or did not (0). The `smoke100` variable indicates whether the 
respondent had smoked at least 100 cigarettes in her lifetime. The other 
variables record the respondent's `height` in inches, `weight` in pounds as well
as their desired weight, `wtdesire`, `age` in years, and `gender`.

###1.  How many cases are there in this data set?  How many variables?  For each variable, identify its data type (e.g. categorical, discrete).

```{r cdc data, eval=TRUE}
str(cdc)
```

There are 20,000 rows which represent cases.

There are 9 variables: genhlth - categorical (ordinal), exerany - categorical, hlthplan - categorical, smoke100 - categorical, height - numeric (discrete), weight - numeric (discrete), wtdesire - numeric (discrete), age - numeric (discrete), and gender - categorical.

## Summaries and tables

The BRFSS questionnaire is a massive trove of information.  A good first step in
any analysis is to distill all of that information into a few summary statistics
and graphics.  As a simple example, the function `summary` returns a numerical 
summary: minimum, first quartile, median, mean, second quartile, and maximum. 


###2.  Create a numerical summary for `height` and `age`, and compute the interquartile range for each. 
For `height` and `age`this is:

```{r summary-weight, eval=TRUE}
summary(cdc$height)
summary(cdc$age)
```

The Interquartile range for height is 

```{r IQR height, eval=TRUE}
70-64
```

The Interquartile range for age is 

```{r IQR age, eval=TRUE}
57-31
```

###Compute the relative frequency distribution for `gender` and `exerany`. How many males are in the sample? What proportion of the sample reports being in excellent health?


```{r   eval=TRUE}
table(cdc$gender)
```

There are 9569 males in the sample.

The frequency distribution for gender and exerany:

```{r   eval=TRUE}
table(cdc$gender)/20000
table(cdc$exerany)/20000
```
The proportion of the sample reporting to be in excellent health:
```{r genhlth excellent, eval=TRUE}
sum(cdc$genhlth == "excellent")/20000*100
```

###3.  What does the mosaic plot reveal about smoking habits and gender?

```{r mosaic-smoke-gender, eval=TRUE}
mosaicplot(table(cdc$gender,cdc$smoke100))
```

More males than females reported smoking at least 100 cigarettes.

## Quantitative data

The following two lines first make a new object called `bmi` and then creates 
box plots of these values, defining groups by the variable `cdc$genhlth`.

```{r boxplot-bmi, eval=TRUE}
bmi <- (cdc$weight / cdc$height^2) * 703
boxplot(bmi ~ cdc$genhlth)
```

###4.  What does this box plot show? 

As BMI increases, the level of health moves progressively from excellent to poor.

###Pick another categorical variable from the data set and see how it relates to BMI. List the variable you chose, why you might think it would have a relationship to BMI,  and indicate what the figure seems to suggest.

```{r boxplot-bmi v , eval=TRUE}
bmi <- (cdc$weight / cdc$height^2) * 703
boxplot(bmi ~ cdc$age)
```

There does seem to be a general trend towards a higher BMI during middle age.


## On Your Own

###-   Make a scatterplot of weight versus desired weight. Describe the relationship between these two variables.

```{r scatterplot of wt v wtdesire, eval=TRUE}
plot(cdc$weight, cdc$wtdesire, pch=".")
```

There is a distinct line that demonstrates that almost no one has a desired weight higher than the one they are at currently.


###-   Let's consider a new variable: the difference between desired weight (`wtdesire`) and current weight (`weight`). Create this new variable by subtracting the two columns in the data frame and assigning them to a new object called `wdiff`.

```{r wdiff, eval=TRUE}
wdiff <- cdc$wtdesire - cdc$weight
str(wdiff)
summary(wdiff)
```

###-   What type of data is `wdiff`? If an observation `wdiff` is 0, what does this mean about the person's weight and desired weight. What if `wdiff` is positive or negative?

Wdiff is a numerical data type, if wdiff is 0 then there is no difference between the desired weight and the current weight. If it is positive then the desired weight is higher than the current weight and a negative wdiff indicates that the desired weight is lower than the current weight.

###-   Describe the distribution of `wdiff` in terms of its center, shape, and spread, including any plots you use. What does this tell us about how people feel about their current weight?

```{r boxplot wdiff, eval=TRUE}
hist(wdiff, breaks = 100)
boxplot(wdiff, ylim = c(-100,100))
```

The histogram shows an almost symmetric but slighty left skewed plot of the data and the boxplot shows that the median is negative. The plots show that there are a greater number of people who want to lose weight from the current weight than those who want to gain weight.

###-   Using numerical summaries and a side-by-side box plot, determine if men tend to view their weight differently than women.

```{r men v women, eval=TRUE}
mweight <- subset(cdc$weight, cdc$gender =="m")
mwtdesire <- subset(cdc$wtdesire, cdc$gender =="m")
mwdiff<- mwtdesire - mweight
summary(mwdiff)
fweight <- subset(cdc$weight, cdc$gender =="f")
fwtdesire <- subset(cdc$wtdesire, cdc$gender =="f")
fwdiff<- fwtdesire - fweight
summary(fwdiff)

boxplot(wdiff ~ cdc$gender, ylim = c(-100,100))

```

More women want to lose weight than men.


###-   Now it's time to get creative. Find the mean and standard deviation of `weight` and determine what proportion of the weights are within one standard deviation of the mean.

```{r mean nad std of wt, eval=TRUE}
summary(cdc$weight)
mean(cdc$weight)
sd(cdc$weight)

within <- ifelse( (cdc$weight > (mean(cdc$weight)-1*sd(cdc$weight))) &
                  (cdc$weight < (mean(cdc$weight)+1*sd(cdc$weight))), 1, 0)
totalSamples <- nrow(cdc)

sum(within) / totalSamples
```

The proportion of the weights that are within one standard deviation of the mean is 70.76%.



